---
title: "Bioconductor Interoperability: Integrating tidybiome with TreeSummarizedExperiment and mia"
author:
  - name: "Tidybiome Authors"
    email: "info@tidybiome.org"
date: "`r Sys.Date()`"
package: tidybiome
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bioconductor Interoperability: Integrating tidybiome with TreeSummarizedExperiment and mia}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Abstract

The Bioconductor ecosystem has coalesced around `SummarizedExperiment` and `TreeSummarizedExperiment` (Huang et al., 2021) as foundational data structures for multi-assay high-throughput genomics and metagenomics. While `TreeSummarizedExperiment` (TreeSE) and downstream analytical packages like `mia` provide rigorous S4 containerization, exploring, wrangling, and visualizing microbiome datasets often requires navigating multi-layered slot accessors that conflict with tidy data principles.

`tidybiome` bridges the gap between Bioconductor S4 rigor and `tidyverse` fluidity. It provides bidirectional, lossless conversion between `tidy_microbiome` and `TreeSummarizedExperiment` objects, enabling researchers to seamlessly interchange between `mia` pipelines and `tidybiome`'s modern compositional tools (Robust CLR, CAFT zero-cell consensus differential abundance, distance-based redundancy analysis, and microbial co-occurrence networks).

---

# 1. Introduction

Bioconductor is the gold standard for reproducible computational biology. In microbiome research:
- **`TreeSummarizedExperiment`** extends `SingleCellExperiment` / `SummarizedExperiment` to store assay matrices alongside row/column hierarchical phylogenetic trees (`rowTree`, `colTree`).
- **`mia`** (Microbiome Analysis) provides an extensive collection of functions operating natively on `TreeSummarizedExperiment` containers.

However, many exploratory analyses, data cleaning steps, and publication visualizations benefit from tidy data representations where samples behave as tibble observations and `dplyr` verbs operate intuitively.

`tidybiome` was built to complement, rather than compete with, the Bioconductor ecosystem.

```
┌────────────────────────────────────────┐             ┌────────────────────────────────────────┐
│     TreeSummarizedExperiment (S4)      │             │        tidy_microbiome (S3)            │
│                                        │  to_tidy    │                                        │
│  • assays: counts, relabundance        │ ──────────> │  • Inherits from tbl_df (sample meta)  │
│  • colData: sample covariates          │             │  • S3 bracket subsetting: tb[i, j]     │
│  • rowData: taxonomic annotations      │ <────────── │  • attr(tb, "assays"): matrices        │
│  • rowTree: phylogenetic ape::phylo    │   to_tse    │  • attr(tb, "tax_table"): tibble       │
│                                        │             │  • attr(tb, "phy_tree"): ape::phylo    │
└────────────────────────────────────────┘             └────────────────────────────────────────┘
```

---

# 2. Installation & Loading

To install `tidybiome` along with core Bioconductor dependencies:

```r
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Bioconductor foundation packages
BiocManager::install(c("SummarizedExperiment", "TreeSummarizedExperiment", "mia"))

# Install tidybiome
devtools::install_github("mladen5000/tidybiome")
```

```{r setup, message=FALSE, warning=FALSE}
library(tidybiome)
library(dplyr)
library(ggplot2)

data("gut_microbiome", package = "tidybiome")
gut_microbiome
```

---

# 3. Two-Way Conversion: `TreeSummarizedExperiment` & `tidybiome`

### 3.1 Exporting to `TreeSummarizedExperiment` (`to_tse`)

When collaborating with researchers whose workflows rely on Bioconductor packages (`mia`, `scater`, `bluster`), you can convert any `tidy_microbiome` object directly into a `TreeSummarizedExperiment`:

```{r to_tse, eval=FALSE}
# Convert to TreeSummarizedExperiment
tse <- to_tse(gut_microbiome)
tse
#> class: TreeSummarizedExperiment 
#> dim: 40 24 
#> metadata(0):
#> assays(1): counts
#> rownames(40): ASV01 ASV02 ... ASV39 ASV40
#> rowData names(3): Kingdom Phylum Genus
#> colnames(24): Sample_01 Sample_02 ... Sample_23 Sample_24
#> colData names(4): treatment diet age depth
#> reducedDimNames(0):
#> mainExpName: NULL
#> altExpNames(0):
#> rowLinks: NULL
#> rowTree: NULL
#> colLinks: NULL
#> colTree: NULL
```

All metadata columns are placed into `colData(tse)`, taxonomy ranks into `rowData(tse)`, and all registered count/transformed assays into `assays(tse)`.

### 3.2 Importing from `TreeSummarizedExperiment` (`as_tidybiome`)

Conversely, when receiving a `TreeSummarizedExperiment` or `SummarizedExperiment` object from a Bioconductor pipeline or public repository (e.g. `curatedMetagenomicData`), convert it into `tidy_microbiome` with a single command:

```{r as_tidybiome, eval=FALSE}
# Convert TreeSE or SummarizedExperiment back to tidy_microbiome
tb_restored <- as_tidybiome(tse)
tb_restored
```

The resulting `tb_restored` is immediately ready for native `dplyr` filtering, compositional modeling, and publication visualizations.

---

# 4. A Hybrid Bioconductor Workflow

Combining Bioconductor's `mia` container storage with `tidybiome`'s statistical tools provides the best of both worlds:

### 4.1 Step 1: Quality Control and Normalization with `tidybiome`
Filter low-depth samples with `dplyr::filter` and apply zero-safe Robust CLR without pseudocount distortion:

```{r rclr_step}
# Tidy filtering and modern rCLR transformation
gut_clean <- gut_microbiome |>
  filter(depth > 12000) |>
  transform_abundance(method = "rclr") |>
  transform_abundance(method = "gmpr")

gut_clean
```

### 4.2 Step 2: Unified Hill Diversity Profiles
Rather than computing disparate index scales (Shannon vs Simpson vs Richness), compute the unified Hill numbers profile ($q = 0, 1, 2$):

```{r alpha_step}
gut_clean <- calc_alpha_diversity(gut_clean, metrics = c("hill", "simplex_variation"))
gut_clean |>
  select(sample_id, treatment, hill_0, hill_1, hill_2, simplex_variation) |>
  head(5)
```

### 4.3 Step 3: Statistical Modeling with Formula Covariate Adjustments
Bioconductor datasets frequently include clinical covariates (age, sex, treatment center). Perform multi-engine consensus differential abundance modeling controlling for confounders:

```{r da_step}
da_res <- calc_differential_abundance(
  gut_clean,
  formula = ~ treatment + age,
  contrast = "treatment",
  methods = c("consensus", "caft", "linda", "clr_linear", "wilcoxon")
)

da_res |>
  filter(is_significant) |>
  select(taxon_id, Phylum, Genus, log2fc, padj_consensus, agreement_score) |>
  head(6)
```

### 4.4 Step 4: Publication Graphics
Produce high-aesthetic visuals without cluttered legends:

```{r viz_step, fig.width=7, fig.height=4}
# Compositional Heatmap with hierarchical clustering and annotation tracks
plot_heatmap(gut_clean, rank = "Genus", top_n = 15, annotation_col = "treatment")
```

---

# 5. Diagnostic Reporting

Generate a standalone, zero-dependency HTML quality control dashboard:

```{r report_step}
tmp_report <- tempfile(fileext = ".html")
report_tidybiome(gut_clean, output = tmp_report, browse = FALSE)
cat("Generated dashboard report at:", tmp_report, "\n")
unlink(tmp_report)
```

---

# 6. Conclusion

`tidybiome` complements the Bioconductor ecosystem by providing:
1. **Lossless Bridges**: Interoperability between `TreeSummarizedExperiment`, `SingleCellExperiment`, `phyloseq`, and `vegan`.
2. **Modern Statistical Parity**: Robust CLR, CAFT zero-cell likelihood ratio modeling, vectorized UniFrac, and distance-based redundancy analysis.
3. **Ergonomic Simplicity**: Native `dplyr` grammar and publication-quality graphics.

---

# 7. References

1. **Huang, R. et al. (2021)**. *TreeSummarizedExperiment: a S4 class for data with hierarchical structure*. **F1000Research**, 9:1246.
2. **Ernst, M. et al. (2025)**. *mia: Microbiome analysis based on the SummarizedExperiment framework*. **Bioconductor**.
3. **Martino, C. et al. (2019)**. *A Novel Sparse Compositional Metric Alleviates Problems in Biomarker Discovery*. **mSystems**, 4(1), e00016-19.
4. **Chen, L. et al. (2018)**. *GMPR: A robust normalization method for zero-inflated count data with application to microbiome sequencing data*. **Microbiome**, 6(1), 215.

---

# 8. Session Information

```{r sessionInfo}
sessionInfo()
```
