---
title: "Getting Started with tidybiome: A Modern, Tidyverse-Native Microbiome Suite"
author: "tidybiome Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tidybiome: A Modern, Tidyverse-Native Microbiome Suite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## 1. Introduction

Downstream microbiome analysis has historically been divided between legacy packages like `phyloseq` and complex Bioconductor frameworks like `mia` / `TreeSummarizedExperiment`. While powerful, these tools often require navigating intricate S4 slots, rely on outdated heuristics (such as arbitrary rarefaction or pseudocount-based CLR), and output visuals with hard-to-read 50-color legends.

**`tidybiome`** brings microbiome data science directly into the modern **tidyverse** ecosystem. It acts as a lightweight, aesthetic Swiss Army knife built around the following core principles:

1. **Native Tidyverse Ergonomics**: The `tidy_microbiome` container inherits from `tbl_df`. Standard `dplyr` verbs (`filter`, `mutate`, `select`, `arrange`, `slice`) work seamlessly on sample metadata while keeping count matrices, taxonomic lineages, distance caches, and phylogenetic trees synchronized under the hood.
2. **State-of-the-Art & bioRxiv Methodologies**:
   * **Robust CLR (rCLR)**: Zero-safe compositional normalization without pseudocount distortion (Martino et al., 2019).
   * **Unified Hill Numbers Profile**: Richness ($q=0$), exponential Shannon ($q=1$), and inverse Simpson ($q=2$) on the intuitive scale of effective number of species.
   * **Phylogenetic UniFrac**: High-performance vectorized unweighted and weighted UniFrac distances via ancestor-descendant matrix algebra.
   * **Consensus Differential Abundance**: Robust multi-engine consensus (LinDA, CAFT zero-cell separation, CLR-linear models, Wilcoxon) with formula specification (`formula = ~ treatment + age + batch`), covariate adjustments, and Cauchy combination testing.
   * **Constrained Ordination & Networks**: Distance-based Redundancy Analysis (db-RDA) and microbial co-occurrence networks with false discovery rate (FDR) control and degree centrality.
3. **Publication-Grade Visualizations**: Soft-gray `"Other"` pooling, colorblind-safe palettes (Okabe-Ito, Nature), confidence ellipses, and `theme_tidybiome()`.

---

## 2. Container Setup and Inspection

Let's load `tidybiome`, `dplyr`, and `ggplot2`, and inspect the built-in benchmark dataset `gut_microbiome` (24 samples across 40 taxa):

```{r setup, message=FALSE, warning=FALSE}
library(tidybiome)
library(dplyr)
library(ggplot2)

data("gut_microbiome", package = "tidybiome")
gut_microbiome
```

Notice the container printout: it displays the sample and feature dimensions, registered assays (`counts`), taxonomy ranks (`Kingdom > Phylum > Genus`), and prints the sample metadata table directly.

---

## 3. Tidy Wrangling and Data Integrity

### 3.1 Using dplyr Verbs
Because `gut_microbiome` is a tibble, you can filter, arrange, and mutate sample metadata directly:

```{r dplyr_verbs}
# Filter to samples with sequencing depth > 15,000 and calculate log10-depth
gut_filtered <- gut_microbiome |>
  filter(depth > 15000) |>
  mutate(log_depth = log10(depth))

gut_filtered
```

All underlying assay matrices (`counts`) are automatically sliced along the column dimension to match the remaining samples!

### 3.2 S3 Bracket Subsetting (`[`, `[<-`)
Direct matrix-style bracket indexing is fully supported:

```{r bracket_subsetting}
# Subset the first 6 samples and select metadata columns
gut_slice <- gut_microbiome[1:6, c("sample_id", "treatment", "age")]
gut_slice
```

### 3.3 Taxonomic Filtering (`filter_taxa`)
To filter along the feature/taxonomic dimension, use `filter_taxa()`. This slices taxonomic rows, assay matrix rows, and automatically prunes phylogenetic tree tips:

```{r filter_taxa}
# Filter to Bacteroidota taxa
gut_bacteroidota <- filter_taxa(gut_microbiome, Phylum == "Bacteroidota")
gut_bacteroidota
```

### 3.4 Container Validation
You can run automated integrity checks at any point:

```{r validation}
validate_tidy_microbiome(gut_bacteroidota, verbose = TRUE)
```

---

## 4. Modern Compositional Transformations (rCLR)

Microbiome sequencing data are high-dimensional, sparse compositions. Adding arbitrary pseudocounts (e.g. $+1$) before logarithmic transformation creates severe artifacts for rare taxa.

`tidybiome` implements **Robust CLR (rCLR)** (Martino et al., 2019), which calculates geometric means restricted strictly to observed non-zero counts:

```{r rclr}
gut_microbiome <- transform_abundance(gut_microbiome, method = "rclr")
gut_microbiome
```

The transformed assay is registered under `rclr` without modifying original sequence counts.

---

## 5. Alpha Diversity: Unified Hill Profiles

Classic alpha diversity metrics (Chao1, Shannon entropy, Gini-Simpson) operate on disparate mathematical scales. `tidybiome` computes the unified **Hill numbers profile**:
* $q = 0$: Species Richness (count of observed taxa)
* $q = 1$: Exponential Shannon entropy (effective number of typical species)
* $q = 2$: Inverse Simpson index (effective number of dominant species)

```{r alpha_diversity}
gut_microbiome <- calc_alpha_diversity(
  gut_microbiome,
  metrics = c("hill", "shannon", "simplex_variation")
)

gut_microbiome |>
  select(sample_id, treatment, hill_0, hill_1, hill_2, simplex_variation) |>
  head(6)
```

---

## 6. Beta Diversity & Phylogenetic UniFrac

### 6.1 Compositional & Ecological Distances
`calc_beta_diversity()` supports:
* `"raitchison"`: Robust Aitchison distance (Euclidean on rCLR-transformed counts).
* `"bray"`: Fast vectorized Bray-Curtis dissimilarity.
* `"jaccard"`: Vectorized binary Jaccard incidence dissimilarity.
* `"tree_wasserstein"`: Earth Mover's Distance across the taxonomic tree hierarchy.

```{r beta_distances}
gut_microbiome <- calc_beta_diversity(gut_microbiome, metric = "raitchison")
```

### 6.2 Phylogenetic UniFrac
When an `ape::phylo` tree is attached (or simulated), `calc_unifrac()` computes unweighted and weighted normalized UniFrac distances via vectorized edge-descendant matrix multiplication:

```{r unifrac}
taxa_ids <- attr(gut_microbiome, "tax_table")$taxon_id
set.seed(42)
mock_tree <- ape::rtree(length(taxa_ids), tip.label = taxa_ids)
gut_phylo <- set_tree(gut_microbiome, mock_tree)

# Compute unweighted UniFrac
u_dist <- calc_unifrac(gut_phylo, weighted = FALSE)
as.matrix(u_dist)[1:4, 1:4]
```

---

## 7. Dimensionality Reduction (RPCA Biplot & cPCA)

### 7.1 Robust PCA (RPCA) Biplots
RPCA decomposes the rCLR assay using singular value decomposition, providing true sample coordinates and taxon loading vectors:

```{r rpca}
gut_microbiome <- calc_ordination(gut_microbiome, method = "rpca")
```

---

## 8. Statistical Rigor: Differential Abundance with Covariates

Microbial differential abundance analysis is notoriously susceptible to false discoveries. `tidybiome` provides a comprehensive **Multi-Engine Consensus Architecture** supporting formula syntax:

```{r da_consensus}
da_results <- calc_differential_abundance(
  gut_microbiome,
  formula = ~ treatment + age,
  contrast = "treatment",
  methods = c("consensus", "caft", "linda", "clr_linear", "wilcoxon")
)

da_results |>
  filter(is_significant) |>
  select(taxon_id, Phylum, Genus, log2fc, padj_consensus, agreement_score) |>
  head(6)
```

* **CAFT** (bioRxiv 2025): Likelihood ratio testing separating zero-cell dropout from abundance shifts.
* **LinDA**: Linear models with rank-based reference taxon selection.
* **CLR-Linear**: Multiple linear regression on centered log-ratio abundances.
* **Cauchy Combination Test**: Aggregates $p$-values across distinct engines without assuming independence.

---

## 9. Constrained Ordination (db-RDA) & Co-Occurrence Networks

### 9.1 Distance-Based Redundancy Analysis (db-RDA)
Test and visualize how environmental metadata variables constrain microbial community variations:

```{r dbrda}
dbrda_model <- run_dbrda(gut_microbiome, formula = ~ treatment + age, distance = "bray")
dbrda_model
```

### 9.2 Microbial Co-Occurrence Networks
Build network graphs with correlation thresholds, FDR adjustment, and node degree centrality:

```{r networks}
net_model <- calc_network(
  gut_microbiome,
  method = "spearman",
  min_prevalence = 0.25,
  r_cutoff = 0.35,
  p_cutoff = 0.05
)

net_model
```

---

## 10. Ecological Screening

Screen communities for prevalence, dominance, baseline divergence, and host-microbe cross-associations:

```{r screening}
# Taxon prevalence and mean abundance
calc_prevalence(gut_microbiome) |> head(4)

# Dominant taxon per sample
calc_dominant(gut_microbiome, add_to_metadata = FALSE)

# Sample divergence relative to Control group
div_res <- calc_divergence(gut_microbiome, reference = list(treatment = "Control"))
mean(div_res$divergence[div_res$treatment == "Control"])
mean(div_res$divergence[div_res$treatment == "Treated"])

# Cross-associations between taxa and clinical variables
calc_cross_association(gut_microbiome, variables = c("age", "depth")) |> head(4)
```

---

## 11. Interoperability with Ecosystem Packages

Convert effortlessly between `tidybiome`, `vegan`, `phyloseq`, and `mia`:

```{r interoperability, eval=FALSE}
# Convert to vegan community matrix and metadata
veg <- to_vegan(gut_microbiome)

# Convert to phyloseq S4 object
ps <- to_phyloseq(gut_microbiome)

# Convert to Bioconductor TreeSummarizedExperiment
tse <- to_tse(gut_microbiome)

# Convert back to tidy_microbiome
tb_from_ps  <- as_tidybiome(ps)
tb_from_tse <- as_tidybiome(tse)
```

---

## 12. Summary

`tidybiome` combines modern compositional statistical rigor, high computational performance, and seamless tidyverse ergonomics into a unified downstream microbiome analysis platform.
