---
title: "Introduction to tidybiome: A Modern, Tidyverse-Native Microbiome Suite"
author: "tidybiome authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tidybiome: A Modern, Tidyverse-Native Microbiome Suite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Introduction

`tidybiome` brings the microbiome data science workflow into the modern **tidyverse** ecosystem. Inspired by Bioconductor's `mia` and legacy `phyloseq`, `tidybiome` acts as a simpler, more aesthetic "Swiss Army knife" that prioritizes state-of-the-art methods from recent literature and bioRxiv preprints:

* **Container Ergonomics**: Objects of class `tidy_microbiome` inherit directly from `tbl_df`, enabling native `dplyr` verbs (`filter`, `mutate`, `select`, `arrange`, `slice`) while keeping count assays, taxonomic tables, and phylogenetic trees synchronized under the hood.
* **Modern Compositional Data Analysis (CoDA)**: Replaces pseudocount heuristics and arbitrary rarefaction with **Robust CLR (rCLR)** and **Robust Aitchison Distance** (Martino et al., 2019).
* **Unified Diversity Profiles**: Calculates Hill numbers profiles ($q=0, 1, 2$) on the intuitive scale of effective species.
* **Dimensionality Reduction**: Implements **Robust PCA (RPCA)** singular value decomposition for compositional biplots with true taxon loading vectors.
* **Consensus Differential Abundance**: Inspired by 2024–2025 bioRxiv preprints (*ConsensusMetaDA*, *dar*, *LinDA*), combines multiple robust engines (LinDA, CLR-linear models, Wilcoxon) with Cauchy combination tests and agreement voting to avoid false discoveries.
* **Publication-Grade Visuals**: Intelligent top-$N$ pooling, colorblind-safe palettes, and minimalist styling via `theme_tidybiome()`.

---

## 2. Quickstart Workflow

```{r setup, message=FALSE, warning=FALSE}
library(tidybiome)
library(dplyr)
library(ggplot2)

# Load the built-in benchmark dataset
data("gut_microbiome", package = "tidybiome")
gut_microbiome
```

### 2.1 Tidy Filtering and Mutation
Because `gut_microbiome` is a tibble on the outside, you can filter and manipulate metadata naturally:

```{r tidy_verbs}
gut_sub <- gut_microbiome |>
  filter(depth > 12000) |>
  mutate(log_depth = log10(depth))
```

### 2.2 Modern Compositional Transformation (rCLR)
```{r rclr}
gut_sub <- transform_abundance(gut_sub, method = "rclr")
```

### 2.3 Alpha Diversity (Unified Hill Numbers)
```{r alpha}
gut_sub <- calc_alpha_diversity(gut_sub, metrics = c("hill", "shannon"))
gut_sub |>
  select(sample_id, treatment, hill_0, hill_1, hill_2) |>
  head(5)
```

### 2.4 Beta Diversity & RPCA Biplot
```{r beta_ord}
gut_sub <- calc_beta_diversity(gut_sub, metric = "raitchison")
gut_sub <- calc_ordination(gut_sub, method = "rpca")
```

### 2.5 Multi-Engine Consensus Differential Abundance
```{r da}
da_res <- calc_differential_abundance(
  gut_sub,
  group = "treatment",
  methods = c("consensus", "linda", "clr_linear", "wilcoxon")
)

# Inspect top biomarkers
da_res |>
  filter(is_significant) |>
  select(taxon_id, Phylum, Genus, log2fc, padj_consensus, agreement_score) |>
  head(6)
```

### 2.6 Publication-Grade Visuals
```{r plots, eval=FALSE}
# Taxonomic composition with intelligent top-N grouping
plot_composition(gut_sub, rank = "Phylum", top_n = 4, group_by = "treatment")

# RPCA Biplot with 95% confidence ellipses and driving taxon vectors
plot_ordination(gut_sub, method = "rpca", color = "treatment", ellipse = TRUE, biplot = TRUE)

# Alpha diversity comparison
plot_alpha(gut_sub, metric = "hill_1", x = "treatment", test = TRUE)

# Biomarker volcano plot
plot_da_volcano(da_res)
```
