---
title: "Statistical Rigor and Ecological Modeling in tidybiome"
author: "tidybiome Authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Rigor and Ecological Modeling in tidybiome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## 1. Overview

Microbiome data present profound statistical challenges:
1. **Extreme Sparsity**: High percentages of true and structural zero counts.
2. **Compositionality**: Sequencing yields relative proportions constrained to the simplex, where shifts in one taxon artificially inflate or deflate others.
3. **High Dimensionality**: Hundreds to thousands of taxa with small sample cohorts ($P \gg N$).
4. **Confounding Covariates**: Age, sex, antibiotic history, batch effects, and sequencing depth bias naive comparisons.

This vignette explains how **`tidybiome`** addresses these challenges using modern statistical methods, formula-based linear modeling, phylogenetic distance vectorization, and ecological modeling.

---

## 2. Zero-Safe Compositional Transformations (Robust CLR)

### 2.1 The Problem with Pseudocounts
The classical Centered Log-Ratio (CLR) transform is defined as:
$$\text{clr}(x_i) = \left[ \log\frac{x_{i1}}{g(x_i)}, \dots, \log\frac{x_{iP}}{g(x_i)} \right]$$
where $g(x_i) = (\prod_{j=1}^P x_{ij})^{1/P}$ is the geometric mean across all taxa.

Because microbiome data contain many zeros, analysts traditionally add an arbitrary pseudocount: $\log(x_{ij} + c)$ for $c \in \{0.5, 1\}$. However, as demonstrated by Martino et al. (2019), the choice of $c$ introduces severe mathematical artifacts:
- Artificially alters sample geometric means.
- Biases sample distance calculations and ordination projections.
- Generates spurious differential abundance biomarkers.

### 2.2 The Robust CLR (rCLR) Solution
`tidybiome` implements **rCLR**, which calculates the geometric mean strictly over observed non-zero counts:
$$g_{>0}(x_i) = \left( \prod_{j: x_{ij} > 0} x_{ij} \right)^{1 / |\{j: x_{ij} > 0\}|}$$
For non-zero taxa:
$$\text{rclr}(x_{ij}) = \log\frac{x_{ij}}{g_{>0}(x_i)}$$
For zero counts ($x_{ij} = 0$), $\text{rclr}(x_{ij})$ remains 0 without any imputation or distortion:

```{r rclr_demo, message=FALSE, warning=FALSE}
library(tidybiome)
data("gut_microbiome", package = "tidybiome")

tb_rclr <- transform_abundance(gut_microbiome, method = "rclr")
# Zero values in raw counts remain exactly zero in rclr
raw_counts <- assay(tb_rclr, "counts")
rclr_counts <- assay(tb_rclr, "rclr")

all(rclr_counts[raw_counts == 0] == 0)
```

---

## 3. Formula-Based Differential Abundance with Covariates

### 3.1 Confounder Control with Formula Syntax
In clinical and ecological surveys, naive two-group tests (like Wilcoxon or two-sample $t$-tests) fail to account for confounding factors like age, BMI, or sequencing batch.

`tidybiome`'s `calc_differential_abundance()` accepts standard R formula syntax:
```r
calc_differential_abundance(tb, formula = ~ treatment + age + batch, contrast = "treatment")
```

### 3.2 Continuous Predictors
In addition to categorical treatments, `tidybiome` supports continuous response modeling (e.g. testing for taxa that correlate with continuous age or biomarker concentration):

```{r continuous_da}
da_continuous <- calc_differential_abundance(
  gut_microbiome,
  formula = ~ age,
  methods = c("clr_linear", "linda")
)

da_continuous |>
  head(5)
```

### 3.3 Engine Architectures
1. **CAFT (bioRxiv Dec 2025)**:
   A two-part likelihood ratio model designed for zero-inflated microbiome data. Separates:
   - Zero-cell dropout probability via logistic regression: $\text{logit}(P(Y=0)) \sim \mathbf{X}\boldsymbol{\beta}$
   - Abundance shifts among present counts via log-linear regression: $\log(Y \mid Y > 0) \sim \mathbf{X}\boldsymbol{\gamma}$
   Combines likelihood ratio test statistics while gracefully suppressing separation warnings.
2. **LinDA (Zhou et al., 2022)**:
   Compositional linear regression selecting a robust reference taxon via mode/rank estimation, fully controlling for model matrix covariates.
3. **CLR-Linear**:
   Fits multiple linear regression models on rCLR or CLR transformed features.
4. **Cauchy Combination Aggregation**:
   Combines $p$-values across disparate statistical frameworks into a single unified consensus $p$-value (`padj_consensus`), weighted by engine agreement score.

---

## 4. Vectorized Phylogenetic Ecology (UniFrac)

### 4.1 Vectorized Descendant-Edge Incidence
Calculating UniFrac traditionally requires deep recursive traversals over the tree for every pair of samples ($O(N^2 \cdot \text{nodes})$), leading many R packages to rely on external C/C++ libraries.

`tidybiome` implements a vectorized algorithm using an **edge-descendant incidence matrix** $\mathbf{M} \in \{0, 1\}^{E \times T}$, where $E$ is the number of edges in the tree and $T$ is the number of tips.
Each tip is traced to the root in a single precomputation step ($O(T \cdot \text{depth})$). Multiplying:
$$\mathbf{W} = \mathbf{M} \times \mathbf{P}$$
where $\mathbf{P} \in [0, 1]^{T \times N}$ is the sample proportion matrix, computes the total descendant branch weight for all edges across all $N$ samples simultaneously in pure R matrix algebra.

### 4.2 Unweighted & Weighted UniFrac
- **Unweighted UniFrac**:
  $$d_{\text{unifrac}}(A, B) = \frac{\sum_e l_e |\mathbb{I}(W_{eA} > 0) - \mathbb{I}(W_{eB} > 0)|}{\sum_e l_e \mathbb{I}(W_{eA} + W_{eB} > 0)}$$
- **Weighted Normalized UniFrac**:
  $$d_{\text{wunifrac}}(A, B) = \frac{\sum_e l_e |W_{eA} - W_{eB}|}{\sum_e l_e (W_{eA} + W_{eB})}$$

```{r unifrac_demo}
taxa_ids <- attr(gut_microbiome, "tax_table")$taxon_id
set.seed(123)
tree <- ape::rtree(length(taxa_ids), tip.label = taxa_ids)
tb_phylo <- set_tree(gut_microbiome, tree)

# Fast vectorized calculations:
dist_unweighted <- calc_unifrac(tb_phylo, weighted = FALSE)
dist_weighted   <- calc_unifrac(tb_phylo, weighted = TRUE, normalized = TRUE)

as.matrix(dist_unweighted)[1:3, 1:3]
as.matrix(dist_weighted)[1:3, 1:3]
```

---

## 5. Distance-Based Redundancy Analysis (db-RDA)

Distance-based Redundancy Analysis (Legendre & Anderson, 1999) is the gold standard for constrained ordination of non-Euclidean ecological distance matrices (Bray-Curtis, Jaccard, UniFrac).

`tidybiome::run_dbrda()` fits the constrained linear ordination:
$$\mathbf{Y}_{\text{dist}} \sim \mathbf{X}_{\text{metadata}}$$

```{r dbrda_demo}
dbrda_fit <- run_dbrda(gut_microbiome, formula = ~ treatment + age, distance = "bray")
dbrda_fit
```

The output returns:
- `samples`: Ordination coordinates for all samples.
- `biplot`: Directional arrows for continuous and categorical predictors.
- `variance_explained`: Proportion of constrained inertia explained by each canonical axis.
- `anova`: Permutation ANOVA table testing overall model significance.

---

## 6. Microbial Co-Occurrence Networks

Pairwise co-occurrence networks identify putative cooperative guilds, mutual exclusions, and ecological keystones:

```{r network_demo}
net <- calc_network(
  gut_microbiome,
  method = "spearman",
  min_prevalence = 0.20,
  r_cutoff = 0.35,
  p_cutoff = 0.05
)

# Nodes table with degree centrality
head(net$nodes, 5)

# Edges table with correlation strength and direction
head(net$edges, 5)
```

The returned `tidybiome_network` object contains tidy data frames that can be directly piped into `plot_network(net)` or external graph packages like `igraph` or `tidygraph`.

---

## 7. Container Integrity & Diagnostic Validation

To ensure analytical reproducibility, `tidybiome` provides `validate_tidy_microbiome()`:

```{r validation_demo}
validate_tidy_microbiome(gut_microbiome, verbose = TRUE)
```

This verifies:
1. **Sample ID Synchronization**: Ensures row names of metadata match column names of all assays.
2. **Taxon ID Synchronization**: Ensures row names of taxonomy match row names of all assays.
3. **Phylogenetic Tip Alignment**: If a tree is present, verifies tree tip labels correspond 1:1 with taxon IDs.

---

## References

1. **Martino, C. et al. (2019)**. *A Novel Sparse Compositional Metric Alleviates Problems in Biomarker Discovery*. **mSystems**, 4(1), e00016-19.
2. **Zhou, H. et al. (2022)**. *LinDA: linear models for differential abundance analysis of microbiome compositional data*. **Genome Biology**, 23, 95.
3. **Lozupone, C. & Knight, R. (2005)**. *UniFrac: a new phylogenetic metric for comparing microbial communities*. **Applied and Environmental Microbiology**, 71(12), 8228-8235.
4. **Legendre, P. & Anderson, M. J. (1999)**. *Distance-based redundancy analysis: testing multispecies responses in multifactorial ecological experiments*. **Ecological Monographs**, 69(1), 1-24.
